---
title: "Verification of Ooshima & Mitamura (2025) 'Effects of perspective-taking training based on relational frame theory for cognitive empathy and emotional empathy'"
subtitle: "Data processing"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

# Dependencies

```{r}

library(tidyverse)
library(readxl)
library(janitor)
library(knitr)
library(kableExtra)

```

# Data processing

NB data loaded is not raw - participants with partial data have already been removed. Raw doesn't seem to be available on the OSF.

Corrispondance with Ooshima revealed that the all_law_data_pre.xlsx and all_law_data_post.xlsx files already have item reverse scoring applied to them, so they are not applied again below.

```{r}

data_pre_iri <- read_xlsx("../data/raw/all_law_data_pre.xlsx", sheet = "IRI law") |>
  rename(id = No) |>
  rename_with(.fn = ~ paste0("iri_", .), .cols = starts_with("Q")) |>
#   # item reverse scoring - not needed as it turns out - Ooshima clarified that the 'raw' data has already been reverse scored
#   mutate(across(
#   c(iri_Q3, iri_Q4, iri_Q7, iri_Q12, iri_Q13, iri_Q14, iri_Q15, iri_Q18, iri_Q19),
#   ~ 6 - .x,
#   .names = "{.col}_rev"
# )) |>
  # mutate(
  #   iri_pt_pre = iri_Q3_rev + iri_Q8 + iri_Q11 + iri_Q15_rev + iri_Q21 + iri_Q25 + iri_Q28,
  #   iri_ec_pre = iri_Q2 + iri_Q4_rev + iri_Q9 + iri_Q14_rev + iri_Q18_rev + iri_Q20 + iri_Q22,
  #   iri_fs_pre = iri_Q1 + iri_Q5 + iri_Q7_rev + iri_Q12_rev + iri_Q16 + iri_Q23 + iri_Q26,
  #   iri_pd_pre = iri_Q6 + iri_Q10 + iri_Q13_rev + iri_Q17 + iri_Q19_rev + iri_Q24 + iri_Q27
  # ) |>
  mutate(
    iri_pt_pre = iri_Q3 + iri_Q8 + iri_Q11 + iri_Q15 + iri_Q21 + iri_Q25 + iri_Q28,
    iri_ec_pre = iri_Q2 + iri_Q4 + iri_Q9 + iri_Q14 + iri_Q18 + iri_Q20 + iri_Q22,
    iri_fs_pre = iri_Q1 + iri_Q5 + iri_Q7 + iri_Q12 + iri_Q16 + iri_Q23 + iri_Q26,
    iri_pd_pre = iri_Q6 + iri_Q10 + iri_Q13 + iri_Q17 + iri_Q19 + iri_Q24 + iri_Q27
  ) |>
  select(id, iri_pt_pre, iri_ec_pre, iri_fs_pre, iri_pd_pre)

data_post_iri <- read_xlsx("../data/raw/all_law_data_post.xlsx", sheet = "IRI law") |>
  rename(id = No) |>
  rename_with(.fn = ~ paste0("iri_", .), .cols = starts_with("Q")) |>
#   # item reverse scoring - not needed as it turns out - Ooshima clarified that the 'raw' data has already been reverse scored
#   mutate(across(
#   c(iri_Q3, iri_Q4, iri_Q7, iri_Q12, iri_Q13, iri_Q14, iri_Q15, iri_Q18, iri_Q19),
#   ~ 6 - .x,
#   .names = "{.col}_rev"
# )) |>
  # mutate(
  #   iri_pt_post = iri_Q3_rev + iri_Q8 + iri_Q11 + iri_Q15_rev + iri_Q21 + iri_Q25 + iri_Q28,
  #   iri_ec_post = iri_Q2 + iri_Q4_rev + iri_Q9 + iri_Q14_rev + iri_Q18_rev + iri_Q20 + iri_Q22,
  #   iri_fs_post = iri_Q1 + iri_Q5 + iri_Q7_rev + iri_Q12_rev + iri_Q16 + iri_Q23 + iri_Q26,
  #   iri_pd_post = iri_Q6 + iri_Q10 + iri_Q13_rev + iri_Q17 + iri_Q19_rev + iri_Q24 + iri_Q27
  # ) |>
  mutate(
    iri_pt_post = iri_Q3 + iri_Q8 + iri_Q11 + iri_Q15 + iri_Q21 + iri_Q25 + iri_Q28,
    iri_ec_post = iri_Q2 + iri_Q4 + iri_Q9 + iri_Q14 + iri_Q18 + iri_Q20 + iri_Q22,
    iri_fs_post = iri_Q1 + iri_Q5 + iri_Q7 + iri_Q12 + iri_Q16 + iri_Q23 + iri_Q26,
    iri_pd_post = iri_Q6 + iri_Q10 + iri_Q13 + iri_Q17 + iri_Q19 + iri_Q24 + iri_Q27
  ) |>
  select(id, iri_pt_post, iri_ec_post, iri_fs_post, iri_pd_post)

# Add MES subscales (Self, Other, Affectedness, Perspective-taking, Imagination)
data_pre_mes <- read_xlsx("../data/raw/all_law_data_pre.xlsx", sheet = "MES law") |>
  rename(id = No) |>
  rename_with(.fn = ~ paste0("mes_", .), .cols = starts_with("Q")) |>
  # # reverse score
  # mutate(across(c(mes_Q3, mes_Q10, mes_Q14, mes_Q15, mes_Q16, mes_Q22), ~6 - .x, .names = "{.col}_rev")) |>
  # # sum scores
  # mutate(
  #   mes_other_pre = mes_Q3_rev + mes_Q7 + mes_Q12 + mes_Q14_rev + mes_Q20,
  #   mes_self_pre = mes_Q6 + mes_Q13 + mes_Q19 + mes_Q21,
  #   mes_affected_pre = mes_Q1 + mes_Q2 + mes_Q8 + mes_Q16_rev + mes_Q22_rev,
  #   mes_perspective_pre = mes_Q4 + mes_Q9 + mes_Q10_rev + mes_Q17 + mes_Q23,
  #   mes_imagination_pre = mes_Q5 + mes_Q11 + mes_Q15_rev + mes_Q18 + mes_Q24
  # ) |>
  # sum scores
  mutate(
    mes_other_pre = mes_Q3 + mes_Q7 + mes_Q12 + mes_Q14 + mes_Q20,
    mes_self_pre = mes_Q6 + mes_Q13 + mes_Q19 + mes_Q21,
    mes_affected_pre = mes_Q1 + mes_Q2 + mes_Q8 + mes_Q16 + mes_Q22,
    mes_perspective_pre = mes_Q4 + mes_Q9 + mes_Q10 + mes_Q17 + mes_Q23,
    mes_imagination_pre = mes_Q5 + mes_Q11 + mes_Q15 + mes_Q18 + mes_Q24
  ) |>
  select(id, mes_other_pre, mes_self_pre, mes_affected_pre, mes_perspective_pre, mes_imagination_pre)

data_post_mes <- read_xlsx("../data/raw/all_law_data_post.xlsx", sheet = "MES law") |>
  rename(id = No) |>
  rename_with(.fn = ~ paste0("mes_", .), .cols = starts_with("Q")) |>
  # # reverse score
  # mutate(across(c(mes_Q3, mes_Q10, mes_Q14, mes_Q15, mes_Q16, mes_Q22), ~6 - .x, .names = "{.col}_rev")) |>
  # # sum scores
  # mutate(
  #   mes_other_post = mes_Q3_rev + mes_Q7 + mes_Q12 + mes_Q14_rev + mes_Q20,
  #   mes_self_post = mes_Q6 + mes_Q13 + mes_Q19 + mes_Q21,
  #   mes_affected_post = mes_Q1 + mes_Q2 + mes_Q8 + mes_Q16_rev + mes_Q22_rev,
  #   mes_perspective_post = mes_Q4 + mes_Q9 + mes_Q10_rev + mes_Q17 + mes_Q23,
  #   mes_imagination_post = mes_Q5 + mes_Q11 + mes_Q15_rev + mes_Q18 + mes_Q24
  # ) |>
  # sum scores
  mutate(
    mes_other_post = mes_Q3 + mes_Q7 + mes_Q12 + mes_Q14 + mes_Q20,
    mes_self_post = mes_Q6 + mes_Q13 + mes_Q19 + mes_Q21,
    mes_affected_post = mes_Q1 + mes_Q2 + mes_Q8 + mes_Q16 + mes_Q22,
    mes_perspective_post = mes_Q4 + mes_Q9 + mes_Q10 + mes_Q17 + mes_Q23,
    mes_imagination_post = mes_Q5 + mes_Q11 + mes_Q15 + mes_Q18 + mes_Q24
  ) |>
  select(id, mes_other_post, mes_self_post, mes_affected_post, mes_perspective_post, mes_imagination_post)

# Add TSSQ subscales (Active, Conceptualization, Perspective, Present moment)
data_pre_tssq <- read_xlsx("../data/raw/all_law_data_pre.xlsx", sheet = "TSSQ law") |>
  rename(id = No) |>
  rename_with(.fn = ~ paste0("tssq_", .), .cols = starts_with("Q")) |>
  mutate(
    tssq_active_pre = tssq_Q3 + tssq_Q9 + tssq_Q10 + tssq_Q14 + tssq_Q16 + tssq_Q17 + tssq_Q20,
    tssq_conceptualization_pre = tssq_Q2 + tssq_Q7 + tssq_Q8 + tssq_Q11 + tssq_Q12 + tssq_Q18,
    tssq_perspective_pre = tssq_Q1 + tssq_Q4 + tssq_Q6,
    tssq_present_pre = tssq_Q5 + tssq_Q13 + tssq_Q15 + tssq_Q19
  ) |>
  select(id, tssq_active_pre, tssq_conceptualization_pre, tssq_perspective_pre, tssq_present_pre)

# Add TSSQ subscales (Active, Conceptualization, Perspective, Present moment)
data_post_tssq <- read_xlsx("../data/raw/all_law_data_post.xlsx", sheet = "TSSQ law") |>
  rename(id = No) |>
  rename_with(.fn = ~ paste0("tssq_", .), .cols = starts_with("Q")) |>
  mutate(
    tssq_active_post = tssq_Q3 + tssq_Q9 + tssq_Q10 + tssq_Q14 + tssq_Q16 + tssq_Q17 + tssq_Q20,
    tssq_conceptualization_post = tssq_Q2 + tssq_Q7 + tssq_Q8 + tssq_Q11 + tssq_Q12 + tssq_Q18,
    tssq_perspective_post = tssq_Q1 + tssq_Q4 + tssq_Q6,
    tssq_present_post = tssq_Q5 + tssq_Q13 + tssq_Q15 + tssq_Q19
  ) |>
  select(id, tssq_active_post, tssq_conceptualization_post, tssq_perspective_post, tssq_present_post)

# Combine into one dataset
data_combined <- data_pre_iri |>
  full_join(data_post_iri, by = "id") |>
  full_join(data_pre_mes, by = "id") |>
  full_join(data_post_mes, by = "id") |>
  full_join(data_pre_tssq, by = "id") |>
  full_join(data_post_tssq, by = "id")

```

# Exclusions

Ooshima clarified by email that their exclusion strategy was as follows:

"Regarding the IQR method: it defines outliers as values falling outside the range between the first quartile (Q1) – 1.5 * IQR and the third quartile (Q3)+ 1.5 * IQR. IQR means Q3 – Q1. As a method, we used Excel's box-and-whisker diagram drawing function to detect each period of each scale, and we also checked the results by manually calculating them."

I.e., IQR ± 1.5IQR, which wasn't explicit in the manuscript.

```{r}

# IQR exclusion strategy function
replace_iqr_outliers_with_na <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  ifelse(x < lower | x > upper, NA, x)
}

data_processed <- data_combined %>%
  mutate(across(-id, replace_iqr_outliers_with_na, .names = "{.col}_outliers_excluded"))

```

# Write to disk

```{r}

write_csv(data_processed, "../data/processed/data_processed.csv")

```

