---
title: "Verification of Ooshima & Mitamura (2025) 'Effects of perspective-taking training based on relational frame theory for cognitive empathy and emotional empathy'"
subtitle: "Comparison of reprocessed data against data stated as having been used in the analysis"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

# Dependencies

```{r}

library(tidyverse)
library(readxl)
library(broom)
library(janitor)
library(effectsize)
library(patchwork)
library(knitr)
library(kableExtra)

```

# Load data

There are three sets of data files present in the supplementary matrials on OSF, which do not seem to agree. I illustrate the point with just the MES perspective data but it may apply to others too.

```{r}

# MES PT
data_mes_pt_1 <- read_xlsx("../data/original/MES_perspective.xlsx") |>
  select(id = No, Gender, Age, pre, post)

data_mes_pt_2 <- read_xlsx("../data/original/import_MESperspective.xlsx") |>
  mutate(Age = as.character(Age)) |>
  select(id = No, Gender, Age, pre = `視点取得_pre`, post = `視点取得_post`)

# IRI PT
data_iri_pt_1 <- read_xlsx("../data/original/IRI_PT.xlsx") |>
  select(id = No, pre, post)

data_iri_pt_2 <- read_xlsx("../data/original/import_total.xlsx") |>
  mutate(Age = as.character(Age)) |>
  select(id = No, pre = PT_pre, post = PT_post)

# TSSQ PT
data_tssq_pt_1 <- read_xlsx("../data/original/TSSQ_perspective.xlsx") |>
  select(id = No, Gender, Age, pre, post)

data_tssq_pt_2 <- read_xlsx("../data/original/import_total.xlsx") |>
  mutate(Age = as.character(Age)) |>
  select(id = No, pre = `視点取り_pre`, post = `視点取り_post`)


# reprocessed
data_reprocessed <- read_csv("../data/processed/data_processed.csv") |>
  # select the data after outlier exclusions and then rename the columns for simplicity 
  select(id, ends_with("_outliers_excluded")) |>
  rename_with(~ str_remove(., "_outliers_excluded$")) |>
  select(id, 
         mes_perspective_pre, mes_perspective_post,
         iri_pt_pre, iri_pt_post,
         tssq_perspective_pre, tssq_perspective_post)

```

# Comparing the analytic dataset vs the other data original file

There are two folders of data, "data" and "data for analysis". By their content and description, both would appear to be 'final' datasets that should in principle be identical. Test whether this is the case: 

## IRI

```{r}

data_iri_pt <- full_join(data_iri_pt_1, 
                         data_iri_pt_2,
                         by = "id") |>
  mutate(match = ifelse(pre.x == pre.y & post.x == post.y, TRUE, FALSE),
         match_if_swapped = ifelse(pre.x == post.y & post.x == pre.y, TRUE, FALSE)) 

data_iri_pt |>
  summarize(proportion_match = mean(match)) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

Pre and post time points are identical in both data files.

## MES

```{r}

data_mes_pt <- full_join(data_mes_pt_1, 
                         data_mes_pt_2,
                         by = c("id", "Gender", "Age")) |>
  mutate(match = ifelse(pre.x == pre.y & post.x == post.y, TRUE, FALSE),
         match_if_swapped = ifelse(pre.x == post.y & post.x == pre.y, TRUE, FALSE)) 

data_mes_pt |>
  summarize(proportion_match = mean(match),
            proportion_match_if_swapped = mean(match_if_swapped)) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

Pre and post time points are swapped in one data file compared to the other. Which one is correct? This would flip the direction of the effect. Further below I recalculate the sum scores from the item level data to check. 

## TSSQ

```{r}

data_tssq_pt <- full_join(data_tssq_pt_1, 
                          data_tssq_pt_2,
                          by = "id") |>
  mutate(match = ifelse(pre.x == pre.y & post.x == post.y, TRUE, FALSE),
         match_if_swapped = ifelse(pre.x == post.y & post.x == pre.y, TRUE, FALSE)) 

data_tssq_pt |>
  summarize(proportion_match = mean(match)) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

Pre and post time points are identical in both data files.

# Comparing the analytic dataset with recalculated sum scores

## IRI

### Pre

```{r}

ggplot(data_iri_pt_2, aes(pre)) +
  scale_x_continuous(limits = c(10, 40)) +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the IRI PT subscale at pre\nfrom the 'for analysis' file")

ggplot(data_reprocessed, aes(iri_pt_pre)) +
  scale_x_continuous(limits = c(10, 40), name = "pre") +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the IRI PT subscale at pre\nreprocessed from the item-level data")

```

### Post

```{r}

ggplot(data_iri_pt_2, aes(post)) +
  scale_x_continuous(limits = c(10, 40)) +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the IRI PT subscale at post\nfrom the 'for analysis' file")

ggplot(data_reprocessed, aes(iri_pt_post)) +
  scale_x_continuous(limits = c(10, 40), name = "post") +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the IRI PT subscale at post\nreprocessed from the item-level data")

```

### Scatter

```{r fig.height=4, fig.width=4}

p_iri_orig <- ggplot(data_iri_pt_2, aes(pre, post)) +
  scale_x_continuous(limits = c(10, 40)) +
  scale_y_continuous(limits = c(10, 40)) +
  #geom_point(alpha = 0.6) +
  geom_jitter(alpha = 0.6, width = .2, height = .2, color = "#35B779FF") +
  theme_linedraw() +
  ggtitle("Data for the IRI PT subscale\nfrom the 'for analysis' file")

p_iri_orig

p_iri_new <- ggplot(data_reprocessed, aes(iri_pt_pre, iri_pt_post)) +
  scale_x_continuous(limits = c(4, 37), name = "pre") +
  scale_y_continuous(limits = c(4, 37), name = "post") +
  #geom_point(alpha = 0.6) +
  geom_jitter(alpha = 0.6, width = .2, height = .2, color = "#31688EFF") +
  theme_linedraw() +
  ggtitle("Data for the IRI PT subscale\nreprocessed from the item-level data")

p_iri_new

```

## MES

### Pre

```{r}

ggplot(data_mes_pt_2, aes(pre)) +
  scale_x_continuous(limits = c(10, 30)) +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the MES PT subscale at pre\nfrom the 'for analysis' file")

ggplot(data_reprocessed, aes(mes_perspective_pre)) +
  scale_x_continuous(limits = c(10, 30), name = "pre") +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the MES PT subscale at pre\nreprocessed from the item-level data")

```

### Post

```{r}

ggplot(data_mes_pt_2, aes(post)) +
  scale_x_continuous(limits = c(10, 30)) +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the MES PT subscale at post\nfrom the 'for analysis' file")

ggplot(data_reprocessed, aes(mes_perspective_post)) +
  scale_x_continuous(limits = c(10, 30), name = "post") +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the MES PT subscale at post\nreprocessed from the item-level data")

```

### Scatter

```{r fig.height=4, fig.width=4}

p_mes_orig <- ggplot(data_mes_pt_2, aes(pre, post)) +
  scale_x_continuous(limits = c(10, 30)) +
  scale_y_continuous(limits = c(10, 30)) +
  #geom_point(alpha = 0.6) +
  geom_jitter(alpha = 0.6, width = .2, height = .2, color = "#35B779FF") +
  theme_linedraw() +
  ggtitle("Data for the MES PT subscale\nfrom the 'for analysis' file")

p_mes_orig

p_mes_new <- ggplot(data_reprocessed, aes(mes_perspective_pre, mes_perspective_post)) +
  scale_x_continuous(limits = c(10, 30), name = "pre") +
  scale_y_continuous(limits = c(10, 30), name = "post") +
  #geom_point(alpha = 0.6) +
  geom_jitter(alpha = 0.6, width = .2, height = .2, color = "#31688EFF") +
  theme_linedraw() +
  ggtitle("Data for the MES PT subscale\nreprocessed from the item-level data")

p_mes_new

```

## TSSQ

### Pre

```{r}

ggplot(data_tssq_pt_2, aes(pre)) +
  scale_x_continuous(limits = c(0, 25)) +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the TSSQ PT subscale at pre\nfrom the 'for analysis' file")

ggplot(data_reprocessed, aes(tssq_perspective_pre)) +
  scale_x_continuous(limits = c(0, 25), name = "pre") +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the MSSQ PT subscale at pre\nreprocessed from the item-level data")

```

### Post

```{r}

ggplot(data_tssq_pt_2, aes(post)) +
  scale_x_continuous(limits = c(0, 25)) +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the TSSQ PT subscale at post\nfrom the 'for analysis' file")

ggplot(data_reprocessed, aes(tssq_perspective_post)) +
  scale_x_continuous(limits = c(0, 25), name = "post") +
  geom_histogram() +
  theme_linedraw() +
  ggtitle("Data for the MSSQ PT subscale at post\nreprocessed from the item-level data")

```

### Scatter

```{r fig.height=4, fig.width=4}

p_tssq_orig <- ggplot(data_tssq_pt_2, aes(pre, post)) +
  scale_x_continuous(limits = c(0, 25)) +
  scale_y_continuous(limits = c(0, 25)) +
  #geom_point(alpha = 0.6) +
  geom_jitter(alpha = 0.6, width = .2, height = .2, color = "#35B779FF") +
  theme_linedraw() +
  ggtitle("Data for the TSSQ PT subscale\nfrom the 'for analysis' file")

p_tssq_orig

p_tssq_new <- ggplot(data_reprocessed, aes(tssq_perspective_pre, tssq_perspective_post)) +
  scale_x_continuous(limits = c(0, 25), name = "pre") +
  scale_y_continuous(limits = c(0, 25), name = "post") +
  #geom_point(alpha = 0.6) +
  geom_jitter(alpha = 0.6, width = .2, height = .2, color = "#31688EFF") +
  theme_linedraw() +
  ggtitle("Data for the TSSQ PT subscale\nreprocessed from the item-level data")

p_tssq_new

```

## Combined scatter plots

```{r fig.height=8, fig.width=8}

p_iri_orig +
  p_iri_new +
  p_mes_orig +
  p_mes_new +
  # p_tssq_orig + # these scores match, but in retrospect they also weren't part of the directional hypotheses. For the preprint, I don't include them in the figure.
  # p_tssq_new +
  plot_layout(ncol = 2)

dir.create("plots")
ggsave("plots/plot_data_comparisons.png", width = 8, height = 8)

```

- IRI and MES produce very different sum scores to that used in the analytic data set

# Session info

```{r}

sessionInfo()

```





